\begin{center}
    \headingfont\huge\bfseries{Abstract}
\end{center}
\begin{changemargin}{1cm}{1cm}
\textit{Today, porting Haskell code to web browsers is complicated. Current approaches such as compilation to JavaScript or WasmMVP suffer from issues such as large binary sizes or performance issues. This is due to semantic differences between the target language and Haskell, as well as having to ship a custom GC in case of WasmMVP. A promising alternative is compiling Haskell to WebAssembly with GC support (WasmGC), which allows compilers to make use of the host environment's GC instead of shipping their own, which can lead to smaller binary sizes and improved maintainability. The goal of this project is to investigate the effects of using WasmGC as a target for Haskell compilers by extending the MicroHs compiler with a new backend and runtime system. Since WasmGC is not a language inteded to be written by hand, we design a custom DSL for generating WasmGC code to help in the implementation of the runtime system. We evaluate the generated Wasm binaries in terms of binary size and runtime performance against the existing MicroHs Wasm backend and native backend. Preliminary results show that the WasmGC backend produces significantly smaller binaries than both the existing Wasm backend and the native backend, and that the WasmGC backend performs significantly better than the existing Wasm backend. We discuss the plan going forward for Part 2 of the thesis, which includes completing the implementation of the WasmGC backend and runtime system, more extensive evaluation of the produced WasmGC binaries, and the development of a demo application to showcase the capabilities of the WasmGC backend.}
\end{changemargin}