\chapter{Compiler and Runtime System Design}\label{ch:runtime}
\section{Compiler and Runtime System Overview}
If you recall the architecture of the MicroHs compiler from Chapter~\ref{ch:background}, the compiler transforms the generated combinators to a useable format for the C runtime, after which it combines the generated code with the runtime system and compiles it to a binary. 

To target WasmGC, we will need to expand on these two steps. Next to generating the C-code for the combinators, we will also generate WasmGC code for constructing the combinator graph. Likewise, next to the C runtime system, we will also implement a runtime system in WasmGC.


\section{Domain Specific Language}
Since WasmGC is a low-level language generally meant to be targeted by compilers instead of being written by hand, writing the runtime system directly in WasmGC would be difficult and error-prone. To solve this issue, we have designed and implemented a simple DSL embedded in Haskell for generating WasmGC code, which abstracts away the explicit stack manipulation that is necessary when writing Wasm code directly. The DSL is an imperative language and consists of three main components: an abstract representation of programs, a code generation function that takes a program in the abstract representation and generates WasmGC code, and a monadic interface for writing programs in the abstract representation.

\subsection{Abstract Representation} 
The abstract representation of programs in the DSL consists of a set of data types that represent the different constructs in the language. The main data type is \texttt{RtMod}, which represents a module in Wasm and contains the type definitions, imports, and functions of the module. The type definitions \texttt{RtTypDef} consist of a name and what type to bind it to, while functions \texttt{RtFunc} have explicitly declared local variables at the function definition level and a body that consists of a list of expressions. Expressions \texttt{RtExpr} can be variable assignments, function calls, control flow constructs such as if statements and while loops, and other constructs such as literals and variable references. The abstract representation allows us to write programs in a more high-level and structured way compared to writing raw WasmGC code, while still being close enough to WasmGC to allow for simple code generation.

\subsection{Code Generation}
For generating the WasmGC code from the AST, we make use of the \texttt{DSL} monad, which has a state that has a generator for generating fresh variable names, settings such as the indent level for pretty printing, and a list of types which we use for simple type checking. We can than generate a wasm program using the function \texttt{$\texttt{toWat} :: \text{String} \rightarrow \text{RtMod} \rightarrow \text{DSL String}$}. This function takes a program represented as a \texttt{String} and a runtime system represented as a \texttt{RtMod} and generates the corresponding WasmGC code as a \texttt{String}. The generated code is in the human-readable WAT format, which can then be compiled to Wasm using a tool such as \texttt{wat2wasm}.

\subsection{Monadic Interface}
The monadic interface allows us to write programs in an imperative style by using \texttt{do} notation while the DSL is embedded in a pure functional language. For this we define two monads, namely \texttt{WasmModM} for defining modules and \texttt{WasmM} for defining functions. The \texttt{WasmModM} monad has three helper functions, these are 
\texttt{$\text{type\_} :: \text{String} \rightarrow \text{RtTyp} \rightarrow \text{WasmModM ()}$} for defining types, \texttt{$\text{import\_} :: \text{String} \rightarrow \text{[RtTyp]} \rightarrow \text{[RtTyp]} \rightarrow \text{WasmModM ()}$} for defining imports, and \texttt{$\text{register} \rightarrow \text{RtFunc} \rightarrow \text{WasmModM ()}$} for registering functions. These functions can than be used as follows to define a module:
\begin{minted}{haskell}
myModule = runWasmModM $ do
    type_ "Node" $
        record [ ("left" , noderef)
               , ("right", noderef) ]
    import_ "add" [i32, i32] [i32] 
    register someFunc -- someFunc is a function defined elsewhere in 
                      -- the code using the WasmM monad
\end{minted}

For defining functions, we make use of the following helper functions: $input$, $output$, and $var$ for defining the input parameters, output parameters, and local variables of a function respectively. Furthermore, we have the $uses$ function for indicating that a function makes use of another function, this returns a partially applied \texttt{Call} constructor that can be used later. Last, we have functions for control flow constructs: $when$ for if statements, $if\_$ for if-else statements, $switch$ for switch statements, and $while$ for while loops. To make the syntax easier to read, we have also defined \texttt{then\_} and \texttt{else\_} functions to be \text{id}, this way we can write code like the following:
\begin{minted}{haskell}
someFunc = runWasmM $ do
    x   <- input "x"   i32
    output i32
    y   <- var   "y"   i32
    add <- uses  "add"

    -- `gets` retrieves the value of a variable, 
    -- `i` is a helper function for creating integer literals
    if_ (gets x .== i 0) $ do
        (then_ $ do
            y =: i 1)
        (else_ $ do
            y =: add [gets x, gets x])
    return_ (gets y)
\end{minted}

\section{WasmGC Backend Implementation}
The abstract representation of expressions in the MicroHs compiler is defined as follows:
\begin{minted}{haskell}
data Exp
    = Var String
    | Lit Lit
    | App Exp Exp
    | Lam Ident Exp
\end{minted}
To generate WasmGC code from this representation, we will need to map the abstract representation to the correct code to construct a graph in our DSL. This is done by the function \texttt{$\text{compileExp} :: \text{Exp} \rightarrow \text{RtExpr}$}, which takes an expression in the abstract representation and generates the corresponding code in our DSL.
\begin{minted}{haskell}
compileExp :: Exp -> RtExpr
compileExp (Var x)         = gets "functions" .!! i tIdx
    where tIdx = read (tail (showIdent v)) :: Int
compileExp (Lit (LPrim p)) = i' $ case primIndex p of
    Just idx -> idx
    Nothing  -> error "Unknown primitive"
compileExp (Lit (Lint n))  = mkValue $ i' n
compileExp (App e1 e2)     = mkNode (compileExp e1) (compileExp e2)
compileExp (Lam _ _)       = error "Lambdas should not exist"
\end{minted}
If we encounter a variable, we generate the instructions to lookup the variable in the \texttt{functions} array (the functions array contains references to nodes representing different initialized and to be initialized functions) and return a reference to the variable in question. In case of a literal that is a primitive or combinator we look up its index in the primitive table and return it as a reference to the index. For other literals such as integers, we generate instructions to box the literal in a \texttt{Value} type. Applications are done by constructing a new node with the two arguments as its children. Lambdas should be eliminated by the time we reach the WasmGC backend, so we can ignore them.

\section{Runtime System Implementation}
The runtime system implements the algorithm discussed in Chapter~\ref{ch:background} for performing graph reduction. The runtime system is implemented in WasmGC using our DSL and consists of multiple types and functions.

\paragraph{Types} For the graph representation, we use a struct named \texttt{Node} with two fields \texttt{left} and \texttt{right} which are references to either other nodes, combinators/primitives, or values (we make use of \texttt{anyref} as it is the common supertype of all references). For the LAS, we use an array of references to nodes, and lastly for values we use a struct named \texttt{Value} with a field \texttt{val} of type \texttt{anyref} to store primitive or fully reduced values. The type definitions for these types are as follows:
\begin{minted}{haskell}
module_ $ do
    type_ "Node" $
        record [ ("left" , anyref)
               , ("right", anyref) ]

    type_ "Stack" $
        array noderef

    type_ "Value" $
        record [ ("val", anyref) ]

    ...

\end{minted}

\paragraph{Functions} For the functions, we have a main function that serves as the entry point for the program, a \texttt{reduce} function that implements the graph reduction algorithm, and a \texttt{step} function that applies the reduction rule for a given combinator or primitive. The main function is responsible for constructing the initial graph and starting the reduction process:
\begin{minted}{haskell}
main $ do
    -- used by the graph construction code
    node      <- var "node" noderef
    n         <- var "n"    i32
    functions <- var "functions" stackref
    -- reduction should return a value
    output valueref
    -- the main function makes use of the reduce function
    reduce    <- uses "reduce"

    -- initialize temporary functions array
    functions =: mkStack (i fs) -- fs is the number of functions in the program
    while (gets n .< i fs) $ do
        -- fill the functions array with empty nodes (I 0)
        setAt (gets functions) (gets n) (mkNode (i' iComb) (i' 0))
        n =: increment (gets n)
    n =: i 0
    -- the graph construction code should be inserted here
    insertProgram
    -- since the program we insert produces a node on the stack,
    -- we need an assignment to the node variable without a right hand side
    node      =: ignore
    -- we free the functions stack here to save some memory
    functions =: nil stackref
    -- start the reduction process
    return_ (reduce [gets node])
\end{minted}
The \texttt{reduce} function implements the graph reduction algorithm as discussed in Chapter~\ref{ch:background}. It initializes an array to be used as the LAS and an index (\texttt{sptr}) to keep track of the top of the LAS. It then enters a loop where there are four cases to consider. Case 1: if the current node is \verb|null|, then we have finished reduction and there was no result. Case 2: if the current node has a left child that is the \verb|I| combinator and a right child of the type value, then we have finished reduction and the result is the value stored in the right child. Case 3: if the current node has a left child that is a combinator or primitive, then we call the \verb|step| function to apply the reduction rule for that combinator or primitive. Case 4: if the current node has a left child that is not a combinator or primitive, then we walk down the left spine of the graph by setting the current node to its left child and pushing it onto the LAS. The code for this is as follows:
\begin{minted}{haskell}
function "reduce" $ do
    root    <- input "root"  noderef
    output valueref
    step    <- uses "step"
    stack   <- var "stack"   stackref
    sptr    <- var "sptr"    i32
    current <- var "current" noderef
    
    -- initialize the LAS, sptr, and current node
    stack   =: mkStack (i 100000) -- we use a fixed maximum size for the LAS
    sptr    =: i (-1)
    current =: gets root
    -- the reduction loop
    while true $ do
        -- Case 1: if current is null, we are done and there was no result
        when (isNull (gets current)) $ do
            return_ (nil valueref)
        -- Case 2: if current.left is an i31ref with value I(8) and current.right 
        --         is of type Value, we have reached a result
        when (getLeft (gets current) `is` i31ref) $ do
            when ((asI32 (getLeft (gets current)) .== i iComb) .&&
                    (getRight (gets current) `is` valueref)) $ do
                return_ (getRight (gets current) `as` valueref)
        -- add current to the LAS and increment sptr
        sptr =: increment (gets sptr)
        setAt (gets stack) (gets sptr) (gets current)
        if_ (getLeft (gets current) `is` i31ref)
        -- Case 3: if current.left is an i31ref we can apply a reduction rule
            (then_ $ do 
            [current, sptr] ==: step [gets current, gets stack, gets sptr])
        -- Case 4: otherwise we continue down the left spine of the graph
            (else_ $ do 
                current =: (getLeft (gets current) `as` noderef))
    -- this return is unreachable but we need it to satisfy the type checker
    return_ (nil valueref)
\end{minted}
The \texttt{step} function has a case for each combinator and primitive, which apply the reduction rule for the combinator/primitive in question. For this is makes use of metavariables to represent arguments extracted from the LAS, this makes the reduction rules look closer to the mathematical definitions of the combinators and primitives. The code for the \texttt{step} function is as follows:
\begin{minted}{haskell}
function "step" $ do
    node  <- input "node"  noderef
    stack <- input "stack" stackref
    sptr  <- input "sptr"  i32
    output noderef
    output i32
    -- the combinator/primitive to reduce
    comb  <- var "comb" i32
    -- the new current node and sptr after reduction
    current <- var "current" noderef
    i_      <- var "i"       i32
    -- the metavariables
    x <- var "x" anyref
    y <- var "y" anyref
    z <- var "z" anyref
    w <- var "w" anyref
    v <- var "v" anyref

    -- extract the combinator/primitive from the node
    comb =: asI32 (getLeft (gets node))
    -- lookup the combinator/primitive and apply the reduction
    switch (gets comb) prims
    -- return the new current node and sptr after reduction
    return_ (Block [ Var current, Var i_ ])
\end{minted}
We make use of the helper function \texttt{prims} that generates the cases for each combinator and primitive based on a primitive table that contains the reduction rules for each combinator and primitive. The reduction rules are defined using a helper function \texttt{redRule} that takes a left node definition, a right node definition, and the amount of arguments to pop from the LAS. Using this helper function, we can easily define reduction rules such as the following rule for the \verb|B| combinator $B \; x \; y \; z = x \; (y \; z)$:
\begin{minted}{haskell}
redRuleB = redRule ln rn n
    where
        ln = x
        rn = mkNode y z
        n  = 3
\end{minted}

\paragraph{Module} Using the types and functions defined above, we can define the complete module for the runtime system as follows:
\begin{minted}{haskell}
module_ $ do
    -- types defined earlier
    ...
    -- we need to import a print function for some primitives
    import_ "print" [i32] []
    -- register the functions defined earlier
    register reduceFunc
    register stepFunc
    -- register the main function as the entry point for the program
    registerMain mainFunc
\end{minted}