\chapter{Preliminary Results}\label{ch:prelim}
Since existing Wasm GCs may not perform well on memory workloads typical of functional programs, we have developed an experiment to quickly determine if our WasmGC port is viable. The simple benchmark program below (inspired by the \texttt|Lambda-Calculus cooked n-ways| benchmark~\cite{lamb-n}) is designed to generate a large number of substitutions.

The program defines a simple Peano number type and implements addition, multiplication, factorial, and equality. It then computes the factorial of 6 (720) and the sum of the first 37 natural numbers (703) plus 17 (720), after which it checks if the two results are equal. The primitive \texttt{pint32} prints an integer to the console, so the program will print 1 if the results are equal and 0 otherwise.

\begin{minted}{haskell}
module Simple where
import qualified Prelude ()
import Primitives

data Peano = Zero | Succ Peano
    deriving ()
data Bool = False | True
    deriving ()

one :: Peano
one = Succ Zero
two :: Peano
two = Succ one
three :: Peano
three = Succ two
add :: Peano -> Peano -> Peano
add Zero     y = y
add (Succ x) y = Succ (add x y)
mul :: Peano -> Peano -> Peano
mul Zero     _ = Zero
mul (Succ x) y = add y (mul x y)
fac :: Peano -> Peano
fac Zero     = Succ Zero
fac (Succ x) = mul (Succ x) (fac x)
eqnat :: Peano -> Peano -> Bool
eqnat Zero     Zero     = True
eqnat Zero     (Succ _) = False
eqnat (Succ _) Zero     = False
eqnat (Succ x) (Succ y) = eqnat x y
sumto :: Peano -> Peano
sumto Zero     = Zero
sumto (Succ x) = add (Succ x) (sumto x)
n5 :: Peano
n5 = add two three
n6 :: Peano
n6 = add three three
n17 :: Peano
n17 = add n6 (add n6 n5)
n37 :: Peano
n37 = Succ (mul n6 n6)
n703 :: Peano
n703 = sumto n37
n720 :: Peano
n720 = fac n6
boolToInt :: Bool -> Int
boolToInt False = 0
boolToInt True  = 1

main :: IO ()
main = (_primitive "pint32") (boolToInt (eqnat n720 (add n703 n17)))
\end{minted}

The results of the benchmark are shown in Table~\ref{tab:benchmark_results}. We compare three different metrics across three different backends: execution time, memory usage, and binary size. The execution time is the minimum execution time out of 1000 runs. Due to the timer used in the native backend, the time always results in 1ms as the timer only has millisecond precision. Memory usage is the resident set size (RSS) of the process after execution and only measured for the Wasm backends.

As can be seen the WasmGC backend performs significantly better than the existing Wasm backend. Comparing the run time to the native backend is not very meaningful due to the low timer resolution, thus we cannot say if the WasmGC backend is faster or slower than the native backend.

The memory usage of the WasmGC backend is significantly lower than the existing Wasm backend. This is likely due to the fact of a memory leak in the module generated by emscripten, as the memory usage is lower when the benchmark is run for a lower amount of runs, while the memory usage of the WasmGC backend is stable across different amounts of runs. 

The binary size of the WasmGC backend is also significantly smaller than the existing Wasm backend and the native backend, this has two main reasons: the WasmGC backend does not need to include a GC implementation in the binary, and the WasmGC backend does not offer support for the full set of primitives the other backends support, which means that the runtime system is significantly smaller.

\begin{table}[t]
    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
        \textbf{Backend} & \textbf{Time (ms)} & \textbf{Memory Usage (KiB)} & \textbf{Binary Size (B)} \\
        \hline
        WasmGC            & 0.991 &  65 640 &   4 574 \\
        Wasm (Emscripten) & 2.250 & 845 232 & 452 264 \\
        Native            & 1.000 &     N/A & 216 184 \\
        \hline
    \end{tabular}
    \caption{Results of the benchmark comparing the WasmGC backend against the existing Wasm backend (through emscripten) and native backend. For the Wasm backends the time is the minimum execution time out of 1000 runs. Memory usage is the resident set size (RSS) of the process after execution and only measured for the Wasm backends.}
    \label{tab:benchmark_results}
\end{table}
